---
title: "MBIO600_Final Rmd_Fertitta_Tramonte_Stark-Kinimaka_Snyder"
author: "Jake Snyder"
date: "2023-12-11"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

Key:
HMS = Hawaiian Monk Seal

```{r packages, include=FALSE}

# Data wrangling and visualization
library(ggplot2)
library(dplyr)

# Bayesian hange point analysis
library(bcp)

```

# Data Wrangling

### Below, the HMS call frequency dataset from Lehua Rock is uploaded and filtered. From this dataset, three subsets are derived: "calls_by_hour_day," "average_calls_per_hour," and "df.summary." "calls_by_hour_day" is utilized to plot Figure 3 and conduct a Kruskal-Wallis test, manual changepoint detection, and bcp() changepoint detection. "average_calls_per_hour" is utilized to plot Figure 3. "df.summary" is utilized to plot Figure 1 and 2.

```{r uploading & filtering call frequency data}

# Upload HMS call frequency data from Lehua Rock
LehuaRock <- read.csv("/Users/gails/Desktop/MonkSealProject/LehuaRock_Analysis_Updated.csv")

# Sort call type to only include HMS calls
LehuaRock <- LehuaRock %>%
  filter(Call %in% c("Croak", "Groan", "Growl", "Moan", "Whoop"))

# Add "Hour" variable to LehuaRock dataset
LehuaRock$Hour <- substr(LehuaRock$Start.time, 1, 2)
LehuaRock$Hour <- gsub(":", "", LehuaRock$Hour)
LehuaRock$Hour <- as.numeric(LehuaRock$Hour)

# Extract dataframe of number of calls per hour for each day
calls_by_hour_day <- table(LehuaRock$Hour, LehuaRock$Date, dnn = c('hour', 'day'))
calls_by_hour_day <- as.data.frame(calls_by_hour_day)
calls_by_hour_day$hour <- as.character(calls_by_hour_day$hour)
calls_by_hour_day$hour <- as.numeric(calls_by_hour_day$hour)

# Define zeros (indicating call frequency before and after deployment) as NA
calls_by_hour_day$Freq[calls_by_hour_day$day == '5/10/2021' 
                       & calls_by_hour_day$hour < 12] <- NA
calls_by_hour_day$Freq[calls_by_hour_day$day == '5/16/2021'
                       & calls_by_hour_day$hour > 15] <- NA

# Filter out NA values
calls_by_hour_day <- calls_by_hour_day %>%
  filter(Freq != "NA")

# Add day vs night label to calls_by_hour_day
is_day <- calls_by_hour_day$hour > 5 & calls_by_hour_day$hour < 19
calls_by_hour_day$day_night <- 'dayvsnight'
calls_by_hour_day$day_night[is_day] <- 'day'
calls_by_hour_day$day_night[!is_day] <- 'night'
head(calls_by_hour_day,10)

# Calculate average number of calls per hour across all days
average_calls_per_hour <- aggregate(Freq ~ hour, mean, data=calls_by_hour_day)
head(average_calls_per_hour,10)

# Create dataframe that includes sd of frequency of calls per hour
df.summary <- calls_by_hour_day %>%
  group_by(hour) %>%
  summarise(
    sd = sd(Freq, na.rm = TRUE),
    Freq = mean(Freq))
head(df.summary,10)

```


# Analysis and Modeling

### Below, a Shapiro-Wilk test for normality is applied to HMS call frequencies and yields a p-value < 0.05, both before and after a sqrt transformation. Thus, the null assuming normality is rejected, and a non-parametric Kruskal-Wallis test is conducted. The Kruskal-Wallis test yields a chi-squared value of 5.6244 and a p-value of 0.01771.

```{r Non-parametric Kruskal-Wallis Test Comparing Day v.s. Night Call Frequencies}

# Plot histogram of calls
hist(calls_by_hour_day$Freq)
# Plot Q-Q plot for calls
qqnorm(calls_by_hour_day$Freq, datax = T)
# Conduct Shapiro-Wilk normality test on non-transformed call frequency data
shapiro.test(calls_by_hour_day$Freq)

# sqrt transform call frequencies
calls_by_hour_day$sqrt_Freq <- sqrt(calls_by_hour_day$Freq)
# Plot histogram of transformed data
hist(calls_by_hour_day$sqrt_Freq)
# Plot Q-Q plot of transformed data
qqnorm(calls_by_hour_day$sqrt_Freq, datax = T)
# Conduct Shapiro-Wilk normality test on transformed call frequency data
shapiro.test(calls_by_hour_day$sqrt_Freq)

# Conduct Kruskal-Wallis test
kruskal.test(calls_by_hour_day$Freq~calls_by_hour_day$day_night)

```

### Below, the hour intervals during which the largest, significant magnitude of change in call frequency occurs are calculated. We can assume that these intervals contain a changepoint in call frequency. The intervals identified are from hours 20-23 (m = 9.444444, p-value = 0.0017497048), hours 9-12 (m = -8.666667, p-value = 0.0003129298), hours 7-10 (m = -8.444444, p-value = 0.0070150348), hours 12-14 (m = 7.285714, p-value = 0.0095862180), and hours 7-12 (m = -6.100000, p-value = 0.0009744568). The interval from hours 9-12 is most significant and has the second-largest slope, suggesting a significant change in call frequency and a potential changepoint.

```{r Identifying Hour Intervals During Which the Largest Magnitude Change in Call Frequency Occurrs (Manual Changepoint Detection)}

HourIntFunc <- function(t0,t2,t3,t4,t5){
  
df <- filter(calls_by_hour_day, hour == t0 | hour == t2 | 
               hour == t3 | hour == t4 | hour == t5)

w.df <- filter(df, hour == t0 | hour == t2)
x.df <- filter(df, hour == t0 | hour == t3)
y.df <- filter(df, hour == t0 | hour == t4)
z.df <- filter(df, hour == t0 | hour == t5)

w.lm <- lm(data = w.df, Freq ~ hour)
x.lm <- lm(data = x.df, Freq ~ hour)
y.lm <- lm(data = y.df, Freq ~ hour)
z.lm <- lm(data = z.df, Freq ~ hour)

w.sum <- summary(w.lm)
x.sum <- summary(x.lm)
y.sum <- summary(y.lm)
z.sum <- summary(z.lm)

w.coef <- w.sum$coefficients[2,c(1,4)]
x.coef <- x.sum$coefficients[2,c(1,4)]
y.coef <- y.sum$coefficients[2,c(1,4)]
z.coef <- z.sum$coefficients[2,c(1,4)]

coef.df <- rbind(w.coef, x.coef, y.coef, z.coef)

strt <- c(t0,t0,t0,t0)
strt <- data.frame(strt)

end <- c(t2,t3,t4,t5)
end <- data.frame(end)

HourInt <- cbind(strt, end, coef.df)

HourInt

}

x0 <-  HourIntFunc(0,2,3,4,5)
x1 <-  HourIntFunc(1,3,4,5,6)
x2 <-  HourIntFunc(2,4,5,6,7)
x3 <-  HourIntFunc(3,5,6,7,8)
x4 <-  HourIntFunc(4,6,7,8,9)
x5 <-  HourIntFunc(5,7,8,9,10)
x6 <-  HourIntFunc(6,8,9,10,11)
x7 <-  HourIntFunc(7,9,10,11,12)
x8 <-  HourIntFunc(8,10,11,12,13)
x9 <-  HourIntFunc(9,11,12,13,14)
x10 <- HourIntFunc(10,12,13,14,15)
x11 <- HourIntFunc(11,13,14,15,16)
x12 <- HourIntFunc(12,14,15,16,17)
x13 <- HourIntFunc(13,15,16,17,18)
x14 <- HourIntFunc(14,16,17,18,19)
x15 <- HourIntFunc(15,17,18,19,20)
x16 <- HourIntFunc(16,18,19,20,21)
x17 <- HourIntFunc(17,19,20,21,22)
x18 <- HourIntFunc(18,20,21,22,23)
x19 <- HourIntFunc(19,21,22,23,0)
x20 <- HourIntFunc(20,22,23,0,1)
x21 <- HourIntFunc(21,23,0,1,2)
x22 <- HourIntFunc(22,0,1,2,3)
x23 <- HourIntFunc(23,1,2,3,4)

HourInt.lm.df <- rbind(x0,x1,x2,x3,x4,x5,x6,x7,x8,x9,x10,
      x11,x12,x13,x14,x15,x16,x17,x18,
      x19,x20,x21,x22,x23)

colnames(HourInt.lm.df) <- c("start", "end", "slope", "p-value")

# Find 10 largest slopes in magnitude
HourInt.lm.df$abs.slope <- abs(HourInt.lm.df$slope)
top10.m <- head(HourInt.lm.df[order(-HourInt.lm.df$abs.slope),], 10)
top10.m

# Identify which slopes are significant (p-value , 0.05)
sig.m <- top10.m[which(top10.m$`p-value` < 0.05),]
sig.m

```

### Below, the hours at which there is a high posterior probability of change in call frequency (i.e. changepoints) are identified using Bayesian change point analysis via the bcp() function. The hours identified are 7, 19, and 22.

```{r Deriving Changepoints via Bayesian change point analysis}

# Plot posterior probability of change at hours
x <- calls_by_hour_day$Freq
bcp_x <- bcp(x, return.mcmc = TRUE)
plot(bcp_x)

# Identify posterior probabilities above 0.25, 0.5, and 0.75
PostProbFunc <- function(data){
x <- data$Freq
bcp_x <- bcp(x, return.mcmc = TRUE)
bcp_sum <- as.data.frame(summary(bcp_x))

bcp_sum$id <- 1:length(x)
sel <- bcp_sum[which(bcp_x$posterior.prob > 0.25), ]
loc <- time(x)[sel$id]
prob25 <- cbind(".25", loc)
prob25 <- data.frame(prob25)
colnames(prob25) <- c("threshold", "loc")

bcp_sum$id <- 1:length(x)
(sel <- bcp_sum[which(bcp_x$posterior.prob > 0.5), ])
loc <- time(x)[sel$id]
prob50 <- cbind("0.5", loc)
prob50 <- data.frame(prob50)
colnames(prob50) <- c("threshold", "loc")

rbind(prob25, prob50)

}

PostProbFunc(calls_by_hour_day)

# Derive hour values that correspond with "loc" (location) values
calls_by_hour_day[c(80,83,116),]

```



# Visualization and Communication of Results

### Figure 1 plots the  call frequencies per hour averaged across all trial days (5/10/21-5/16/21). Grey areas indicate calls occuring during nightime. Light areas indicate calls occurring during daytime.

```{r Figure 1. Average Call Frequencies Per Hour - plot on slide 10}

AvgCallF_daynight <- df.summary %>%
  ggplot(aes(x = hour, y = Freq)) +
  geom_rect(aes(xmin = -Inf, xmax = 6, ymin = -Inf, ymax = Inf), 
            fill = "lightgray", alpha = 0.4) +
  geom_rect(aes(xmin = 19, xmax = Inf, ymin = -Inf, ymax = Inf), 
            fill = "lightgray", alpha = 0.4) +
  geom_line(size = 1.2, color = "black") +
  geom_point(size = 3, shape = 21, fill = "black") +
  geom_errorbar(aes(ymin = Freq-sd, ymax = Freq+sd), 
            width = 0.2, color = "black") +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, face="bold", size=14, color="black"),
    axis.title.x = element_text(face="bold", size=12, color="black"),
    axis.title.y = element_text(face="bold", size=12, color="black"),
    panel.grid.major = element_line(linetype = "solid", color = "grey"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.margin = margin(20, 20, 20, 20)) +
  labs(title = "Average Calls Per Hour", y = "Average Number of Calls", x = "Hour")

AvgCallF_daynight

```

### Figure 2 plots the call frequencies per hour averaged across all trial days (5/10/21-5/16/21). Grey areas indicate calls occuring during nightime. Light areas indicate calls occurring during daytime. Hour intervals during Which the largest magnitude change in call frequency occurrs are indicated in purple (hours 20-23), green (hours 9-12), orange (hours 7-10), blue (hours 12-14), and pink (hours 7-12). Red dots indicate the changepoints derived via the bcp() function (hours 7, 19, and 22).

```{Figure 2. r Average Call Frequencies Per Hour with Manual Changepoint Hour Ranges and bcp()-derived Changepoints Identified - plot on slide 13}

segment_strtend <- data.frame(x1 = 20, x2 = 23, x3 = 9, x4 = 12, x5 = 7, x6 = 10, x7 = 12, x8 = 14, x9 = 7, x10 = 12, y1 = 9.666658, y2 = 37.99999, y3 = 35, y4 = 8.999996, y5 = 39.5, y6 = 14.16667, y7 = 8.999997, y8 = 23.57142, y9 = 39.5, y10 = 9)

AvgCallsPerHour_chngpt <- ggplot(df.summary, aes(x = hour, y = Freq)) +
    geom_rect(aes(xmin = -Inf, xmax = 6, ymin = -Inf, ymax = Inf), 
        fill = "grey", alpha = 0.05) +
    geom_rect(aes(xmin = 19, xmax = Inf, ymin = -Inf, ymax = Inf), 
        fill = "grey", alpha = 0.05) +
    geom_line(size = 1.2, color = "black") + 
    geom_point(size = 3, shape = 21, fill = "black") +
    geom_errorbar(aes(ymin = Freq-sd, ymax = Freq+sd), width = 0.4) +
    labs(title = "Average Calls Per Hour", y = "Average Number of Calls", 
        x = "Hour") +
    # manually-derived changepoint hour ranges defined 
    # & color-coded using geom_segment()
    geom_segment(aes(x = x1, y = y1, xend = x2, yend = y2), 
        colour = "purple", linewidth = 1.5, data = segment_strtend) + 
    geom_segment(aes(x = x3, y = y3, xend = x4, yend = y4), 
        colour = "darkgreen", linewidth = 1.5, data = segment_strtend) + 
    geom_segment(aes(x = x5, y = y5, xend = x6, yend = y6),
        colour = "orange", linewidth = 1.5, data = segment_strtend) +   
    geom_segment(aes(x = x7, y = y7, xend = x8, yend = y8),
        colour = "blue", linewidth = 1.5, data = segment_strtend) + 
    geom_segment(aes(x = x9, y = y9, xend = x10, yend = y10), 
        colour = "hotpink", linewidth = 1.5, data = segment_strtend) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5, 
            face="bold", size=14, color="black"),
          axis.title.x = element_text(face="bold", size=12, color="black"),
          axis.title.y = element_text(face="bold", size=12, color="black"),
          panel.grid.major = element_line(linetype = "solid", 
            color = "darkgrey", linewidth = .01),
          panel.grid.minor = element_blank(),
          panel.border = element_blank(),
          panel.background = element_blank(),
          plot.margin = margin(10, 10, 10, 10)) +
   geom_point(data = df.summary[c(20,23,8),], aes(x = hour, y = Freq), 
        colour="red", size = 5) # bcp()-derived changepoints in red

AvgCallsPerHour_chngpt

```

### Figure 3 plots the call frequencies per hour with each day (5/10/21-5/16/21) of data collection plotted seperately.

```{r Figure 3. Call Frequencies Per Hour with Each Day of Data Collection Plotted Seperately - plot on slide 14}

CallsPerHour_alldays <- 
    ggplot(data = calls_by_hour_day, mapping= aes(x = hour, y = Freq)) +
    geom_rect(aes(xmin = -Inf, xmax = 6, ymin = -Inf, ymax = Inf), 
        fill = "grey", alpha = 0.05) +
    geom_rect(aes(xmin = 19, xmax = Inf, ymin = -Inf, ymax = Inf), 
        fill = "grey", alpha = 0.05) +
    geom_line(aes(color = day), linetype= "longdash", alpha = .4, linewidth = .8) + 
    scale_color_brewer(palette = "Dark2") +
    geom_point(data = average_calls_per_hour, aes(x = hour, y = Freq), 
            size = 3, shape = 21, fill = "black") +
    geom_line(data = average_calls_per_hour, aes(x=hour, y=Freq, 
            color = "average"), linewidth = 1, color = "black") + 
    labs(title = "Calls Per Hour", y = "Number of Calls", x = "Hour") +
    theme(legend.position = "bottom") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5, 
            face="bold", size=14, color="black"),
            axis.title.x = element_text(face="bold", size=12, color="black"),
            axis.title.y = element_text(face="bold", size=12, color="black"),
            panel.grid.major = element_line(linetype = "solid", 
            color = "darkgrey", linewidth = .01),
            panel.grid.minor = element_blank(),
            panel.border = element_blank(),
            panel.background = element_blank(),
            plot.margin = margin(10, 10, 10, 10))

CallsPerHour_alldays

```

